Combining language resources into a grammar-driven Swedish parser

### 2011-10-24 doc/paper/Paper.tex
newer paper

\begin{abstract}
This paper describes on-going work of a rule-based, open-source
parser for Swedish. The central component is a wide-coverage grammar
implemented in the Grammatical Framework formalism (GF), from which
we obtain a %can be compiled to a % from which we derive?? 
parser.
In addition to GF, we use two other main resources: the Swedish
treebank Talbanken and the electronic SALDO lexicon.
The resulting parser will 
give a full syntactic analysis of the input sentences.
It will be highly reusable, freely available,
and as GF provides libraries for compiling
grammars to a number of programming languages,
chosen parts of the the 
grammar may be used in various NLP applications. % dealing with
%controlled or free
%natural language. 
% allow easy interaction with other programs and grammars
%>> something very commercial, interesting and outstanding here please!

@@

This paper will briefly introduce GF, Talbanken and SALDO in Section \ref{sec:background}
and describe our results so far in Section \ref{sec:progress}.
%We aim to implement % The goal is to be able to 
%a parser for open domain language and 
%This paper describes our work and results 
%so far as well as the directions of future work.

@@

The grammar acts as an independent module.
%and may be used in different projects.
Reusability is further supported by the separation between resource grammars
and application grammars. The resource library provided with GF 
implements morphological and
syntactical rules for more than 20 different languages.
Hence the writer of an application grammar can start her work at a higher
level and does not need to describe how to 
form standard sentences, phrases or inflect words.

Since many of the languages in the GF library resemble each other grammatically,
they can share much of their implementations. This is usually done by using a
\verb|Functor|. % , which lets a number of languages share parts of their implementation. 
This avoids code duplication and aids the code maintenance.

@@

\subsection{Talbanken}
Talbanken\cite{talbanken} is
a Swedish treebank assembled in the 1970s at Lund University.
In 2005 it was modernized by Nivre, Nilsson and Hall\cite{talbanken05} and
enriched with annotation for a full phrase structure analysis. 
We are using the treebank for development and evaluation.

Although Talbanken contains both written and spoken Swedish,
only the prose material, consisting of 6316 sentences, will be used in this project.
This part was also used when training the data-driven parser Maltparser \cite{malt}. \\
%and the parsetree generated by our grammar will eventually be 
%compared to those from the treebank.

\subsection{Saldo}
SALDO\cite{saldo} is an open source lexicon resource
based on Svenskt Associationslexikon. It is
developed at Språkbanken at Gothenburg University
and intended for usage in language technology
research. 
A large GF lexicon has been extracted from SALDO,
containing 50 000 entries. By updating this technique, we will
re-import SALDO, which has been developed since.
The importing method should be fast and reliable
enough to allow us to always have a fresh version of the dictionary
in GF.

%compare!!
\section{Related work}
Language technology for Swedish is an area of much interesting research.
There are two other deep grammar-based parser for Swedish:
SICS (Raynen and Gambäck, 1992) based on Core Language Engine and a
unification-based chart parser (Mats Wirén, 1988),
though none of them are freely available.
Among other parsers for Swedish, the statistical MaltParser\cite{malt}
which was trained on Talbanken, is worth mentioning. 
CassSwe (Kokkinakis and Johansson-Kokkinakis, 1999) is based on finite state cascades,
whereas the shallow parser GTA (Knutsson, Bigert and Kann, 2003) relies on rules of 
a context free grammar. Both GTA and CassSwe operate on POS tagged text.\\
Extract and FM\cite{MarkusForsberg2007} are tools for supervised lexicon
extraction compatible with GF.
%For lexical extractions, there are tools like
%Extract for supervised lexicon extraction and
%FM for programming lexical resources.

@@

We have so far concentrated on shorter sentences, without idioms and
conjunction. 
Since GF demands information 
that is not given by the Talbanken tags -- such as the valency of verbs -- a
full translation requires that all words used are already known to GF. 
The evaluation of trees of this sort shows that the mapping
can restore 85\% of the nodes. 
%, given that all words are
%known to the grammar.
If the we lift the restrictions, we can currently restore 63\% of the tags.
%mapping, a number that would probably increase substansially when working with a bigger lexicon.
%is an interesting way of
%% an straightforward
%accomplishing an evaluation.

@@

\subsection{Development of the grammar}
For Swedish, about 85\% of the GF resource code is shared with the other Scandinavian
languages. % >> When trying to achieve a lightweight, usable code? Not
%due to the resemblances between the languages, they share
%>> using a functor for sharing code, for when having the same structure but
%>> but only different words. 
However, if we aim for a deeper and more comprehensive analysis of Swedish,
the implementation of the languages needs to be more independent.
%<<in order to avoid ad-hoc solutions and 
%Some constructions that are standard in
%Norwegian and Danish may be very rare in Swedish and while extending the
%grammar there may occur situations where a deeper restructuring of the
%implementation is needed. \\
The resource grammar gives a good start and our present grammar 
covers constructions such as declarative sentences with normal or inverted
word order, questions, passives, imperatives, relative clauses, cleft
constructions etc.
A number of constructs that are generally not present in other languages and
therefore not given by the resources, have also been added. These include the use of
the reflexive pronoun \emph{sitt}: \\
\emph{Han såg \textbf{sitt} hus} $\; \; \; \;$ (\emph{He saw his (own) house}) $\;$
as opposed to \\
\emph{Han såg \textbf{hans} hus} $\;$ (\emph{He saw his (another person's) house}) \\
%>> not critical, does not have to be in resource grammar
%>> also not expressable in all languages
Fronting words or phrases is very common in Swedish and are now allowed by the grammar:\\
%the possibility for this has been added to the grammar:\\
\emph{\textbf{Glad} var han inte.} $\;$ (\emph{\textbf{Happy} was he not}). \\
This sort of rephrasing is not given by the resource grammar, since it
has little effect on the logical representation.

@@

%% more explaination of probabilities
We aim to make the parser robust by equipping it with techniques such as chunk
parsing, named entity recognition and methods for handling unknown 
grammatical constructions such as idioms and ellipses. %etc. 
It is already possible to add probabilities for GF functions,
which rank the parse trees.
By adding dependency probabilities, we aim to improve the
potential for disambiguation.

>>>>

\begin{abstract}
This paper describes on-going work of a rule-based, open-source
parser for Swedish. The central component is a wide-coverage grammar
implemented in the Grammatical Framework formalism (GF), from which
we obtain a %can be compiled to a % from which we derive?? 
parser.
In addition to GF, we use two other main resources: the Swedish
treebank Talbanken and the electronic SALDO lexicon.
The resulting parser 
gives a full syntactic analysis of the input sentences.
It will be highly reusable, freely available,
and as GF provides libraries for compiling
grammars to a number of programming languages,
chosen parts of the the 
grammar may be used in various NLP applications. % dealing with
%controlled or free
%natural language. 
% allow easy interaction with other programs and grammars
%>> something very commercial, interesting and outstanding here please!

@@

This paper will briefly introduce GF, Talbanken and SALDO in Section \ref{sec:background}
and describe our results in Section \ref{sec:progress}.
%We aim to implement % The goal is to be able to 
%a parser for open domain language and 
%This paper describes our work and results 
%so far as well as the directions of future work.

@@

The grammar acts as an independent module and
%and may be used in different projects.
reusability is further supported by the separation between resource grammars
and application grammars. The resource library provided with GF 
implements morphological and
syntactical rules for more than 20 different languages.
Hence the writer of an application grammar can start her work at a higher
level and does not need to describe how to 
form standard sentences, phrases or inflect words.

Since many of the languages in the GF library resemble each other grammatically,
they can share much of their implementations. This is usually done by using a
\verb|Functor|, % , which lets a number of languages share parts of their implementation. 
avoids code duplication and aids the code maintenance.

@@

\subsection{Talbanken}
For development and evaluation, we use the Swedish treebank
Talbanken\cite{talbanken}.
It was assembled in the 1970s at Lund University and modernized
in 2005 by Nivre, Nilsson and Hall\cite{talbanken05} and
enriched with annotation for a full phrase structure analysis.  \\
Although Talbanken contains both written and spoken Swedish,
only the prose material, consisting of 6316 sentences, will be used in this project.
This part was also used when training the data-driven parser Maltparser \cite{malt}. \\
%and the parsetree generated by our grammar will eventually be 
%compared to those from the treebank.

\subsection{Saldo}
SALDO\cite{saldo} is an open source lexicon resource
based on Svenskt Associationslexikon. It is
developed at Språkbanken at Gothenburg University
and intended for usage in language technology
research. 
We are developing tools for extracting GF lexicons
from SALDO, which have resulted in a dictionary containing
over 98000 entries including information about verbs that
are reflexive or needs a particle.
The importing method should be fast and reliable
enough to allow us to always have a fresh version of the dictionary
in GF.

%compare!!
\section{Related work}
Language technology for Swedish is an area of much interesting research.
There are two other deep grammar-based parser for Swedish:
Swedish Core Language Engine (Raynen and Gambäck, 1992) and a
unification-based chart parser\cite{wiren},
%correct reference?
though none of them are freely available.
Among other parsers for Swedish, the statistical MaltParser\cite{malt},
trained on Talbanken, is worth mentioning. 
CassSwe (Kokkinakis and Johansson-Kokkinakis, 1999) is based on finite state cascades,
whereas the shallow parser GTA (Knutsson, Bigert and Kann, 2003) relies on rules of 
a context free grammar. Both GTA and CassSwe operate on POS tagged text.\\
Extract and FM\cite{MarkusForsberg2007} are tools for supervised lexicon
extraction compatible with GF.
%For lexical extractions, there are tools like
%Extract for supervised lexicon extraction and
%FM for programming lexical resources.

@@

We have so far concentrated on shorter sentences, without idioms and
conjunction. 
Since GF demands information 
that is not given by the Talbanken tags -- such as the valency of verbs -- a
full translation requires that all words used are already known to GF. 
The evaluation of trees of this sort shows that the mapping
can restore 85\% of the nodes. 
%, given that all words are
%known to the grammar.
If we lift the restrictions, we can restore 63\% of the tags.
%mapping, a number that would probably increase substantially when working with a bigger lexicon.
%is an interesting way of
%% an straightforward
%accomplishing an evaluation.

@@

\subsection{Development of the grammar}
For Swedish, about 85\% of the GF resource code is shared with the other Scandinavian
languages. % >> When trying to achieve a lightweight, usable code? Not
%due to the resemblances between the languages, they share
%>> using a functor for sharing code, for when having the same structure but
%>> but only different words. 
However, if we aim for a deeper and more comprehensive analysis of Swedish,
the implementation of the languages needs to be more independent.
%<<in order to avoid ad-hoc solutions and 
%Some constructions that are standard in
%Norwegian and Danish may be very rare in Swedish and while extending the
%grammar there may occur situations where a deeper restructuring of the
%implementation is needed. \\
The resource grammar gives a good start and our present grammar 
covers constructions such as declarative sentences with normal or inverted
word order, questions, passives, imperatives, relative clauses, cleft
constructions etc.
A number of constructs that are generally not present in other languages and
therefore not given by the resources, have also been added. These include the use of
the reflexive pronoun \emph{sitt}: \\
\emph{Han såg \textbf{sitt} hus} $\; \; \; \;$ (\emph{He saw his (own) house}) $\;$
as opposed to \\
\emph{Han såg \textbf{hans} hus} $\;$ (\emph{He saw his (another person's) house}) \\
%>> not critical, does not have to be in resource grammar
%>> also not expressible in all languages
Fronting words or phrases is very common in Swedish and are now allowed by the grammar:\\
%the possibility for this has been added to the grammar:\\
\emph{\textbf{Glad} var han inte.} $\;$ (\emph{\textbf{Happy} was he not}). \\
This sort of rephrasing is not given by the resource grammar, since it
has little effect on the logical representation.

@@

%% more explanation of probabilities
We aim to make the parser robust by equipping it with techniques such as chunk
parsing, named entity recognition and methods for handling unknown 
grammatical constructions such as idioms and ellipses. %etc. 
It is already possible to add probabilities for GF functions,
which rank the parse trees.
By adding dependency probabilities, we aim to improve the
potential for disambiguation.
